Experience shapes individual foraging specialization and success in a virtual predator-prey system

### 2023-10-25 manuscript/Chapter2-manuscript.Rmd
arranged spacing in doc
update intro and knit

The capacity of predators to match their tactic to the behaviour of their prey is expected to drive the outcome of predator-prey interactions. To do so, predators need to practice their tactics extensively to develop the skills and expertise to successfully hunt their prey, which may result in individual foraging specialization. Yet, there are very limited empirical assessments showing links between experience, foraging specialization, and performance at the individual level, due to the numerous challenges of monitoring direct interactions in the wild. In this study, we used a virtual predator-prey system (the online game *Dead by Daylight*) to investigate if and how hunting experience shapes individual foraging specialization and success in predators across repeated interactions with their prey. Our results show that predators optimized prey consumption as they gained experience. Prey behaviour was an important mediator of this relationship, which also drove individual differences in the development of expertise. With experience, differences among individuals in foraging specialization increased, as individual predators became either specialist or flexible foragers. Yet, many individuals remained specialists throughout the study period. Moreover, the matching between the predator's strategies (i.e. flexible or specialized foraging) and the prey was not strong, suggesting that part of the population did not match their tactic to the prey. This individual heterogeneity coincided with differences among individuals in success through time. While flexible foragers performed slightly less than specialist foragers, both achieved similar success overall. Our observations suggest that individual differences in behavioural change with experience may drive among individual differences in hunting tactics, which may act as a stabilizing mechanism promoting behavioural diversification in predator-prey systems.

@@

Predator express a diverse array of hunting tactics to capture their prey. In natural populations, differences among individuals in tactics such as slow and fast movement speeds or high attack frequencies are associated with differences in prey consumption [@toscanoTraitmediatedFunctionalResponses2014;@Patrick.Weimerskirch2014;@toscanoPersonalityForagingBehavior2016]. Prey usually respond to predation cues by reducing their activity, or use rapid or unpredictable movements to successfully escape when they are chased [@Walker.etal2005;@Kelley.Magurran2011;@Herbert-Read.etal2017]. This results in reciprocal interactions where predators fine-tune their tactics to their prey, suggesting that they learn and become efficient hunters through repeated encounters [@Estes.etal2003;@Woo.etal2008;@Phillips.etal2017]. Ultimately, the consequences of experience on the outcome of predator-prey interactions could have larger implications for ecosystem dynamics [@Tinker.etal2009;@Dukas2019;@wooster2023].

Empirical studies on human and non-human hunters show that individuals optimize foraging efficiency (e.g. search and handling times, return rates) by learning associative images or prey and environmental cues [@Edwards.Jackson1994;@Morse2000;@MacDonald2007;@Reid.etal2010;@Wilson-Rankin2015]. As predators practice and refine their tactics, they acquire expertise (e.g. skills, knowledge) that allows them to outperform novice hunters [@Dukas2019;@wooster2023]. Through this process, it is hypothesized that individuals will specialize in a hunting tactic (e.g. always moving fast) if it allows them to maintain successful prey captures at each atempt [@Tinker.etal2009;@Dukas2019]. When predators learn to hunt in environments where resources are predictable (or stable), they should gradually specialize in their hunting tactics with experience because individuals require less energy and time to search for, capture, and handle prey [@Weimerskirch2007;@Woo.etal2008;@Potier.etal2015]. These conditions favor specialist hunters because there would be higher costs to switch tactics for the same type of prey. An alternative hypothesis is that learning is a mechanism driving behavioural flexibility. For instance, predators can gain expertise and information on their prey through repeated hunting atempts, and may hunt optimally by learning to adjust their tactic to the type of prey that they encounter [@Stephens1993;@Ishii.Shimada2010;@Kelley.Magurran2011]. While there are costs to attempt novel tactics by trial and error in stable environments, predators that hunt in environments under fluctuating resources can reduce the consequences of uncertainty by learning to switch hunting tactics [@dall2010]. Such feedbacks between ecological interactions and learning make it challenging to predict under which circumstances predators should specialize with experience [@Stephens1993;@Ishii.Shimada2010;@vilapouca2021;@beyts2023].

There is increasing evidence suggesting that the emergence of behavioural specialization and its fitness consequences is driven by resource fluctuation [@Patrick.Weimerskirch2014a;@vandenBosch.etal2019;@Santoro.etal2019;@Manlick.etal2021]. If predators in different environments adjust their hunting to local prey fluctuations, specialist and flexible hunters can both attain similar hunting success and coexist [@Phillips.etal2017]. For example, @Szopa-Comley.Ioannou2022 showed that predatory blue acaras (*Andinoacara pulcher*) reduce their maximum approach speed when repeatedly exposed to prey that escape unpredictably, enabling them to be as successful as those exposed to predictable prey. Therefore, if specialist and flexible hunters can attain equal success by learning through repeated interactions with their prey, then empiricists should investigate the link between foraging specialization and success across time and contexts.

A recurring challenge impeding research on predator-prey behavioural interactions, at the individual level, is the need to collect data simultaneously on both the predator and the prey. Individual-based models can mitigate some of these challenges and have contributed in stirring solid testable ideas for empiricists [Refs]. Yet, while providing maximum control over environmental parameters, they can suffer from emerging behaviours that are unrealistic for animals in nature [refs]. Videogames offer an interesting alternative to agent-based simulations, providing controled virtual environments where real players can interact. For example, in the predator-prey videogame *Dead by Daylight*, four prey players need to forage for resources while avoiding predation by a fifth player. In this virtual system, the predators' hunting tactics and success are driven by the movement of the prey, and the prey can increase their chances of survival by cooperating and moving fast when they forage [@Cere.etal2021;@FraserFranco.etal2022, Santostefano et al. in press]. Predators should thus benefit from learning how the prey behave to be successful. Of course, the life of neither the predator nor the prey players are at stake, such that emerging patterns could be driven by their motivation to win and not "true" survival. Yet, the game often ellicits natural reactions in players such as fear from predation (personal observations), which corroborates with another virtual ecological study showing that predation drives individual variation in risk perception [@Beauchamp2020]. In the game Age of Empire II, @lymberyetal.2023 also showed that complex environments favor strong soldiers over large armies in the game as well as in the laboratory with two ant species. These examples outline how general ecological patterns can emerge from interactions in virtual systems with fixed rules, which in turn might help to advance predator-prey research. Videogames generate large volumes of data on interacting players throughout their lifetime in the game under realistic, controlled, and repeatable ecological contexts. Hence, they offer the opportunity to tackle fundamental questions about the role of experience on predator-prey interactions.

In this study, we assess how repeated experience shapes predator hunting specialization and success using data from players in *Dead by Daylight*. First, we investigate how individual hunting success changes with experience (i.e. how individuals develop their hunting expertise). We hypothesize that the predators' success will increase with experience up to a certain level where it will stagnate. We expect this pattern to vary among individuals due to differences in prey encounters. For example, a predator may face greater difficulty than other individuals if it encountered elusive groups of prey more often. Second, we assess how foraging specialization emerges with experience.

We define hunting tactics as the mean movement speed of the predator along a slow-fast continuum, and the level of individual foraging specialization as the intra-individual variance (i.e. IIV) in movement speed across matches, where low IIV translates to individual specialization (e.g. always using a fast tactic).

If experience reduces the costs of switching between low and high speeds as predators encounter varying groups of prey, we predict that individuals (and thus the population) should all converge towards flexible speeds. However, if individuals encounter similar groups of prey, then the efficiency of a specific tactic should increase with experience and everyone should instead specialize in similar speeds. In both scenarios, among individual variation in IIV (i.e. individual differences in specialization) would be low across experience, whereas the population variance would either increase (flexibility) or decrease (specialization). Alternatively, both processes may operate at the same time if behavioural interactions with the prey vary among predators, such that among individual variation in IIV would increase. In this case, we expect predators that experienced more predictable encounters with their prey (i.e. similar groups across matches) to specialize in similar speeds, while predators that experienced unpredictable encounters with their prey (i.e. heterogeneous groups across matches) should adopt flexible movement speeds. If we detect such prey-dependent fine-tuning, then all hunters along the flexible-specialist continuum should attain equal success probably because they learned from experience how to adjust their hunting style to their prey.

>>>>

The capacity of predators to match their tactic to the behaviour of their prey is expected to drive the outcome of predator-prey interactions. To do so, predators need to practice their tactics extensively to develop the skills and expertise to successfully hunt their prey, which may result in individual foraging specialization. Yet, there are very limited empirical assessments showing links between experience, foraging specialization, and hunting success at the individual level, due to the numerous challenges of monitoring direct interactions in the wild. In this study, we used a virtual predator-prey system (the online game *Dead by Daylight*) to investigate if and how hunting experience shapes individual foraging specialization and success in predators across repeated interactions with their prey. Our results show that predators optimized prey consumption as they gained experience. Prey behaviour was an important mediator of this relationship, resulting in differences among predators in the development of expertise. Throughout experience, 68% of predators became either increasingly specialized by always moving at a fast pace, or flexible by swithing between slow and fast speeds. The remaining 32% of predators used a specialized tactic of fast movement throughout the study period. Contrary to our predictions, the predator's strategies (i.e. flexible or specialized foraging) was not strongly matched to the preys' speed, suggesting that changes in hunting behaviour were driven more by repeated experience than the prey's speed. Ultimately, flexible foragers were only slightly less successful than specialist foragers, such that both achieved similar success overall. Our observations suggest that experience drives among individual differences in hunting tactics, which may act as a stabilizing mechanism promoting behavioural diversification in predator-prey systems.

@@

Predator express a diverse array of hunting tactics to capture their prey. In natural populations, differences among individuals in tactics such as ambushing, fast movement speeds, or high attack frequencies are associated with differences in prey consumption [@toscanoTraitmediatedFunctionalResponses2014;@Patrick.Weimerskirch2014;@toscanoPersonalityForagingBehavior2016]. Prey usually respond to predation cues by reducing their activity, or use rapid or unpredictable movements to successfully escape when they are chased [@Walker.etal2005;@Kelley.Magurran2011;@Herbert-Read.etal2017]. This results in reciprocal interactions where predators fine-tune their tactics to their prey, suggesting that they learn and become efficient hunters through repeated encounters [@Estes.etal2003;@Woo.etal2008;@Phillips.etal2017]. These behavioural adjustments with experience can shape the outcome of predator-prey interactions, which may result in larger implications for community and ecosystem dynamics [@wooster2023].

Empirical studies on human and non-human hunters show that individuals optimize foraging efficiency (e.g. search and handling times, return rates) by learning associative images or prey and environmental cues [@Edwards.Jackson1994;@Morse2000;@MacDonald2007;@Reid.etal2010;@Wilson-Rankin2015]. As predators practice and refine their tactics, they acquire expertise (e.g. skills, knowledge) that allows them to outperform novice hunters [@Dukas2019;@wooster2023]. Through this process, it is hypothesized that individuals will specialize in a hunting tactic (e.g. always moving fast) if it allows them to maintain successful prey captures at each atempt [@Tinker.etal2009;@Dukas2019]. When predators learn to hunt in environments where resources are predictable (or stable), they should gradually specialize in their hunting tactics with experience because individuals require less energy and time to search for, capture, and handle prey [@Weimerskirch2007;@Woo.etal2008;@Potier.etal2015]. These conditions favor specialist hunters because there would be higher costs to switch tactics for the same type of prey. An alternative hypothesis is that learning is a mechanism driving behavioural flexibility. For instance, predators can gain expertise and information on their prey through repeated hunting atempts, and may hunt optimally by learning to adjust their tactic to the type of prey that they encounter [@Stephens1993;@Ishii.Shimada2010;@Kelley.Magurran2011]. While there are costs to attempt novel tactics by trial and error in stable environments, predators that hunt in environments under fluctuating resources can reduce the consequences of uncertainty by learning to switch hunting tactics [@dall2010;@mathotAdaptiveStrategiesManaging2012]. Such feedbacks between ecological interactions and learning make it challenging to predict under which circumstances predators should specialize with experience [@Stephens1993;@Ishii.Shimada2010;@vilapouca2021;@beyts2023].

There is increasing evidence suggesting that the emergence of behavioural specialization and its fitness consequences is driven by resource fluctuations [@Patrick.Weimerskirch2014a;@vandenBosch.etal2019;@Santoro.etal2019;@Manlick.etal2021]. If predators that hunt in different environments adjust their tactics to local prey fluctuations, then specialist and flexible hunters can both attain similar hunting success and coexist [@Phillips.etal2017]. For example, @Szopa-Comley.Ioannou2022 showed that predatory blue acaras (*Andinoacara pulcher*) reduce their maximum approach speed when repeatedly exposed to prey that escape unpredictably, enabling them to be as successful as those exposed to predictable prey. Therefore, if specialist and flexible hunters can attain equal success by learning through repeated interactions with their prey, then empiricists should investigate the link between foraging specialization and success across time and contexts.

A recurring challenge impeding research on predator-prey behavioural interactions is the need to collect data simultaneously on both the predator and the prey. Individual-based models can mitigate some of these challenges and have contributed in stirring testable ideas for empiricists [Refs]. Yet, while providing maximum control over environmental parameters, they can suffer from emerging behaviours that are unrealistic for animals in nature [refs]. Videogames offer an interesting alternative to agent-based simulations, providing controled virtual environments where real players interact. For example, in the predator-prey videogame *Dead by Daylight*, four prey players need to forage for resources while avoiding predation by a fifth player. In this virtual system, the predator population is composed of individuals that either ambush or hunt at high speeds (i.e. mean movement speed along a slow-fast continuum), and their success is driven by the movement of the prey [@FraserFranco.etal2022]. We found that prey can increase their chances of survival by cooperating and moving fast when they forage [@Cere.etal2021;@, Santostefano et al. in press]. Of course, the life of neither the predator nor the prey players are at stake, such that emerging patterns could be driven by their motivation to win and not "true" survival. Yet, the game ellicits natural reactions in players such as fear from predation (personal observations), which corroborates with another virtual ecological study showing that predation drives individual variation in risk perception [@Beauchamp2020]. In a different system, @lymberyetal.2023 showed that complex environments favor strong soldiers over large armies in a virtual game as well as in the laboratory with two ant species. These observations outline how ecological patterns can emerge from interactions in virtual systems with fixed rules. Videogames also generate large volumes of data on interacting players throughout their lifetime in the game under realistic, controlled, and repeatable ecological scenarios. Hence, virtual systems such as *Dead by Daylight* offer the opportunity to tackle fundamental questions about the role of experience on predator-prey interactions.

In this study, we assess how repeated experience shapes predator hunting specialization and success using data from players in *Dead by Daylight*. First, we investigate how individual hunting success changes with experience (i.e. how individuals develop their hunting expertise). We hypothesize that the predators' success will increase with experience up to a certain level where it will stagnate. We expect this pattern to vary among individuals due to differences in prey encounters. For example, a predator may face greater difficulty than other individuals if it encountered elusive groups of prey more often. Second, we assess how foraging specialization emerges with experience. We define the level of individual foraging specialization as the intra-individual variance (i.e. IIV) in movement speed across matches, where low IIV translates to individual specialization (i.e. always using a cursorial tactic). If all predators encounter varying groups of prey, we predict that individuals (and thus the population) should all converge towards flexible speeds. However, if individuals encounter similar groups of prey, then everyone should instead specialize in similar speeds. In both scenarios, among individual variation in IIV (i.e. individual differences in specialization) should be low across experience, whereas the population variance would either increase (flexibility) or decrease (specialization). Alternatively, individual specialization and flexibility may emerge if individuals experience different interactions with their prey, such that among individual variation in IIV would increase. In this case, we expect predators that experienced more predictable encounters with their prey (i.e. similar groups across matches) to specialize in similar speeds, while predators that experienced unpredictable encounters with their prey (i.e. heterogeneous groups across matches) should adopt flexible movement speeds. If we detect such prey-dependent fine-tuning, then all hunters along the flexible-specialist continuum should attain equal success if they learned from experience how to adjust their hunting style to their prey.
